Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-23 paper/ms.tex
a few more edits

Many astronomical objects have characteristic timescales determined by their
physical structure.
Variability is excited by noisy physical processes, and grows most strongly
at the characteristic timescale, but is also damped due to dissipation in the
system.
As a concrete example, stellar (asteroseismic) oscillations are strong at
resonant frequencies determined by the internal stellar structure, which are
both excited and damped by convective turbulence.

@@

For these definitions, the kernel is
\begin{equation}\eqlabel{sho-kernel}
k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
\begin{cases}
    \cosh{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
    2\,(1+\omega_0\,\tau), & Q = 1/2\\
    \cos{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
We note that, because of the damping, the characteristic oscillation frequency
in this model $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
the frequency of the undamped oscillator $\omega_0$.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celerite\ process.
This means that the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example, the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
that for large $\omega$ falls off exponentially.
When compared to \eq{celerite-psd}~--~that for large $\omega$ goes as
$\omega^{-4}$, at most~--~it is clear that a \celerite\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celerite\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celerite\ model are indistinguishable
from the inferences made using the correct model but substantially less
computation time is required for the \celerite\ inference.

@@

In this section we demonstrate several use cases of \celerite\ when
applied to real datasets.
Each of these examples touches on an active area of research so we limit
our examples to be qualitative in nature and make no claim of optimality but
we hope that these examples encourage interested readers to investigate
the applicability of \celerite\ to their research.

@@

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890 observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ mission.
For KIC~11615890, the published asteroseismic parameters measured using several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}
We randomly select a month-long segment of \kepler\ data, initialize our
\celerite\ model using a grid search in the parameter space, and then use
\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
the full set of parameters.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Aigrain:2015} and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the period of the oscillation.
The key difference between this function and other quasiperiodic kernels is
that it is everywhere positive.
We can construct a simple \celerite\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

To demonstrate this method, we fit a \celerite\ model with a kernel given by
\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58$ measured using
traditional periodogram and autocorrelation function approaches applied to
\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
We used \emcee\ to sample the joint posterior probability density for the four
parameters in \eq{rot-kernel} conditioned on two quarters of \kepler\ data and
find a constraint on the rotation period of $P = \rotationperiod$ in good
agreement with the literature value.
\Figure{rotation} shows a subset of the data used in this example and the
posterior inferences of the PSD and autocorrelation function of the process.
\Figure{rotation-period} shows the marginalized posterior distribution for the
rotation period.

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
Like \celerite, the likelihood function for a CARMA model conditioned on a
sorted one-dimensional dataset can be solved in $\mathcal{O}(N)$ using a
recursive method.
The kernel function for CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celerite\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not true, however, for any $J > 1$.
This means that \celerite\ could be used to compute any CARMA model but using
CARMA to compute the likelihood of a \celerite\ model would require solving
\eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
This is an interesting distinction because, while the computational scaling of
CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
somewhat faster and the memory requirements are smaller.
However, as discussed in \sect{sho}, \celerite\ models have physical a
interpretation as a mixture of stochastically-driven damped harmonic
oscillators.
This enabled, for example, the use of \celerite\ as a model for asteroseismic
oscillations.
A similar example using CARMA would necessarily be qualitative.

>>>>

Many astronomical objects are variable on timescales determined by their
physical structure.
For phenomena such as stellar (asteroseismic) oscillations, variability is
excited by noisy physical processes and grows most strongly at the
characteristic timescale but is also damped due to dissipation in the system.
These oscillations are strong at resonant frequencies determined by the
internal stellar structure, which are both excited and damped by convective
turbulence.

@@

For these definitions, the kernel is
\begin{equation}\eqlabel{sho-kernel}
k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
\begin{cases}
    \cosh{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
    2\,(1+\omega_0\,\tau), & Q = 1/2\\
    \cos{(\eta\,\omega_0\,\tau)} +
        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
We note that, because of the damping, the characteristic oscillation frequency
in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
the frequency of the undamped oscillator, $\omega_0$.

@@

For this examplewe simulate a dataset using a known GP model with a kernel
outside of the support of a \celerite\ process.
This means the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

For this example the data are simulated from a quasiperiodic GP with the
kernel
\begin{eqnarray}\eqlabel{sim-wrong-true}
k_\mathrm{true} (\tau) = \alpha\,
    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
\end{eqnarray}
where $P_\mathrm{true}$ is the fundamental period of the process.
This autocorrelation structure corresponds to the power spectrum
\begin{eqnarray}
S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
\right]
\end{eqnarray}
which, for large values of $\omega$, falls off exponentially.
When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
$\omega^{-4}$ at most~--~it is clear that a \celerite\ model can never
perfectly reproduce the structure of this process.
That being said, we demonstrate that rigorous inferences can be made
about $P_\mathrm{true}$ even with an effective model.
The left panel of \Figure{simulated-wrong} shows the simulated dataset.
We then fit this simulated data using the product of two SHO terms
(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
We note that using \eq{product-rule}, the product of two \celerite\ terms can
also be expressed using \celerite.
We apply log-uniform priors on all the parameters and use \emcee\ to sample
the joint posterior probability for all of the parameters.
The inferred distribution for the parameter $P$ is shown in the right panel of
\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
the inferences made using the correct model (\eqalt{sim-wrong-true}).
The inference made using this effective \celerite\ model are indistinguishable
from the inferences made using the correct model but substantially less
computation time is required for the \celerite\ inference.

@@

In this section we demonstrate several use cases of \celerite\ when
applied to real datasets.
Each of these examples touches on an active area of research so we limit our
examples to be qualitative in nature and do not claim that \celerite\ is the
optimum method but we hope these examples encourage interested readers to
investigate the applicability of \celerite\ to their research.

@@

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890, observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ mission.
For KIC~11615890, the published asteroseismic parameters measured using several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}
We randomly select a month-long segment of \kepler\ data, initialize our
\celerite\ model using a grid search in the parameter space, and then use
\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
the full set of parameters.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
\citep{Dumusque:2014}.
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star \citep{Aigrain:2015} and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the period of the oscillation.
The key difference between this function and other quasiperiodic kernels is
that it is positive everywhere.
We can construct a simple \celerite\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

To demonstrate this method, we fit a \celerite\ model with a kernel given by
\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
This star has a published rotation period of $3.88 \pm 0.58$, measured using
traditional periodogram and autocorrelation function approaches applied to
\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
We used \emcee\ to sample the joint posterior probability density for the four
parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
and find a constraint on the rotation period of $P = \rotationperiod$.
This is in good agreement with the literature value.
\Figure{rotation} shows a subset of the data used in this example and the
posterior inferences of the PSD and autocorrelation function of the process.
\Figure{rotation-period} shows the marginalized posterior distribution for the
rotation period.

@@

The continuous autoregressive moving average (CARMA) models introduced into
the astrophysics literature by \citet{Kelly:2014} share many features with
\celerite.
Like \celerite, the likelihood function for a CARMA model conditioned on a
sorted one-dimensional dataset can be solved in $\mathcal{O}(N)$ using a
recursive method.
The kernel function for CARMA$(J,\,K)$ model is \citep{Kelly:2014}
\begin{eqnarray}\eqlabel{carma-kernel}
k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{carma-coeff}
A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
\end{eqnarray}
and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
the model.
Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
every CARMA model corresponds to an equivalent \celerite\ model and the
parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
The inverse statement is not true however, for any $J > 1$.
This means \celerite\ could be used to compute any CARMA model but using
CARMA to compute the likelihood of a \celerite\ model would require solving
\eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
This is an interesting distinction because, while the computational scaling of
CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
somewhat faster and the memory requirements are smaller.
However, as discussed in \sect{sho}, \celerite\ models have a physical
interpretation as a mixture of stochastically-driven damped harmonic
oscillators.
This enables, for example, the use of \celerite\ as a model for asteroseismic
oscillations.
A similar example using CARMA would necessarily be qualitative.
